---
title: "Project 1 - ST595"
author: "Ruslan Mamedov"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
always_allow_html: yes
subtitle: Inferential vs. Predictive Analysis for 2015 Public Use Microdata Sample
  (PUMS)
---

## Introduction

We are given a set of of anonymized census data and asked to conduct two types of statistical analysis to answer the following questions:

1. Inferential analysis:
Do people living in houses pay more on electricity than those living in apartments? How much? Make sure you adjust for (at least) the number of bedrooms and number of occupants in the household.

2. Predictive analysis:
Create a model that could be used to predict electricity costs for a household in Oregon. 

We are then asked to compare and contrast the approaches across the two tasks.

## Data Description

We are provided with the .csv file 'or_acs_house.csv' which contains household level responses to the American Community Survey for households in Oregon. The dataset is a Public Use Microdata Sample (PUMS) from the 2015 1-year survey obtained from http://www2.census.gov/programs-surveys/acs/data/pums/2015/1-Year/. It contains only households that have at least one person, pay for their electricity, and are not group accommodation, and it may be assumed this is a random sample of all such households in Oregon.


## Statistical Modeling

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(include=FALSE)
library(broom)
library(knitr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(plyr) 
library(leaps)
library(caTools)
library(NLP)
library(nlme)
library(dplyr)
library(boot)
library(formatR)
library(ggthemes)
library(DAAG)
library(Metrics)
library(sjmisc)
```

1. Explanatory Problem: Do people living in houses pay more on electricity than those living in apartments, after accounting for the confounding variables? If so, by how much?


H`~0`: The mean electricity payments for people living in houses is not greater than for those living in the apartments, after controlling for differences in the other proposed explanatory variables. 

H`~A`: The mean electricity payments for people living in houses is more than for those living in the apartments after controlling for differences in the other proposed explanatory variables.


Because this is a question of statistical inference, we will fit a model to the data, test the model to see if it satisfies the linear model assumptions, and then interpret it by determining the statistical significance of the coefficients for our proposed explanatory variables. We will determine p-value for the explanatory variable of interest. 

\renewcommand{\arraystretch}{2}
```{r, echo = FALSE, include=TRUE}
PUMS_2015 <- read.csv('OR_acs_house_occ.csv')
PUMS_2015.nice<-PUMS_2015
colnames(PUMS_2015.nice)<-c("Serial number","Records per housing record","Type of unit","Lot size","Number of bedrooms","Units in structure", "Electricity (monthly cost)", "Yearly fuel cost (excl. gas and electricity)","Gas (monthly cost)", "House heating fuel","Number of Rooms", "Tenure", "Property value", "When structure first built", "Under 18 years in household", "60 years and over in household")
kable(head(PUMS_2015.nice, 2), format="latex", booktabs=TRUE) %>% 
  kable_styling(latex_options=c("scale_down","HOLD_position") , font_size = 5, full_width = TRUE)
```

```{r}
str(PUMS_2015)
summary(PUMS_2015)
```

\renewcommand{\arraystretch}{2}
```{r }
PUMS_2015.int = as.data.frame(PUMS_2015.nice[,1])
PUMS_2015.chr = as.data.frame(PUMS_2015.nice[,1])
for (i in 1:ncol(PUMS_2015.nice)){
  if (!is.numeric(PUMS_2015.nice[,i])){
      PUMS_2015.chr<-cbind(PUMS_2015.chr,PUMS_2015.nice[,i])
      colnames(PUMS_2015.chr)[ncol(PUMS_2015.chr)]<-colnames(PUMS_2015.nice[i])
  }
  else{
      PUMS_2015.int<-cbind(PUMS_2015.int,PUMS_2015.nice[,i])
      colnames(PUMS_2015.int)[ncol(PUMS_2015.int)]<-colnames(PUMS_2015.nice[i])
  }
}
PUMS_2015.int <- subset (PUMS_2015.int, select = -c(1))
PUMS_2015.chr <- subset (PUMS_2015.chr, select = -c(1))
```

We have quite a large dataset with over 15000 observations and a mix of numerical and categorical values. First, let's have some exploratory analysis done: spread of electricity bills around different building types, distribution of unique values for both numerical and categorical columns, and summary of missing values.   
```{r , fig.width=6, fig.height=3,echo = FALSE, include=TRUE, warning=FALSE, message=FALSE}
#Boxplots on electricity bills for different building types
ggplot(data=PUMS_2015, aes(factor(BLD), ELEP, fill = factor(BLD))) + 
  geom_boxplot()+theme_tufte()+theme(text = element_text(size=rel(3)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+xlab(colnames(PUMS_2015.nice[6]))+ylab(colnames(PUMS_2015.nice[7]))
```

```{r , fig.width=6, fig.height=4,echo = FALSE, include=TRUE, warning=FALSE, message=FALSE}
#summary of non-numeric column values:
#knitr::kable(list(as.data.frame(table(PUMS_2015.chr[,1])),as.data.frame(table(PUMS_2015.chr[,2])),as.data.frame(table(PUMS_2015.chr[,3])),as.data.frame(table(PUMS_2015.chr[,4])),as.data.frame(table(PUMS_2015.chr[,5])),as.data.frame(table(PUMS_2015.chr[,6])),as.data.frame(table(PUMS_2015.chr[,7]))))%>% kable_styling(latex_options = c("striped","HOLD_position"), stripe_color = "gray!15", font_size = 2.5, position = "left")
library(colorspace)
p1<-ggplot(as.data.frame(table(PUMS_2015.chr[,1])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[1]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_sequential("Purple-Yel")
p2<-ggplot(as.data.frame(table(PUMS_2015.chr[,2])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[2]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_sequential("Reds 2")
p3<-ggplot(as.data.frame(table(PUMS_2015.chr[,3])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[3]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_sequential("Heat")
p4<-ggplot(as.data.frame(table(PUMS_2015.chr[,4])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[4]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_sequential("Plasma")
p5<-ggplot(as.data.frame(table(PUMS_2015.chr[,5])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[5]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")
p6<-ggplot(as.data.frame(table(PUMS_2015.chr[,6])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[6]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_diverging("Cyan-Magenta")
p7<-ggplot(as.data.frame(table(PUMS_2015.chr[,7])), aes(Var1, Freq, fill = Var1))+geom_col()+xlab(colnames(PUMS_2015.chr[7]))+ylab("Frequency")+theme_tufte()+theme(text = element_text(size=rel(2)),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none")+scale_fill_discrete_diverging("Blue-Red2")

#summary of numeric columns
#knitr::kable(summary(PUMS_2015.int))%>% kable_styling(latex_options="scale_down", font_size = 4, full_width = TRUE)
p8<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,2])) + geom_histogram()+xlab(colnames(PUMS_2015.int[2]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p9<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,3])) + geom_bar(width =0.1)+xlab(colnames(PUMS_2015.int[3]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))+xlim(0,2)
p10<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,4])) + geom_histogram()+xlab(colnames(PUMS_2015.int[4]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p11<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,5])) + geom_histogram()+xlab(colnames(PUMS_2015.int[5]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p12<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,6])) + geom_histogram()+xlab(colnames(PUMS_2015.int[6]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p13<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,7])) + geom_histogram()+xlab(colnames(PUMS_2015.int[7]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p14<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,8])) + geom_histogram()+xlab(colnames(PUMS_2015.int[8]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))
p15<-ggplot(PUMS_2015.int, aes(x=PUMS_2015.int[,9])) + geom_histogram()+xlab(colnames(PUMS_2015.int[9]))+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(2)))


grid.arrange(p1,p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13, p14,p15, nrow = 3)

#N/A values
PUMS_2015.na<-PUMS_2015.nice %>% summarise_all(~ sum(is.na(.)))
kable(PUMS_2015.na)%>%kable_styling(latex_options=c("scale_down","HOLD_position"), font_size = 5, full_width = TRUE)
```



Some conclusions:
1. Most of the records are for one-family detached houses.  
2. Number of bedrooms/rooms have a normal distribution.  
3. Property value and Electricity monthly cost have right-skewed disptributions.  
4. Type of unit has min=max=1. The dataset includes only housing units and not group quarters which we shall account for when interpreting the results.   
5. Electricity monthly cost and couple of other columns seem to have outliers (e.g. ELEP=540) not associated with other variables and most likely representing some cut-off value arbitrarily chosen for very large electric bills. Outliers  in electricity column are pretty evenly distributed between housing vs. apartment units, small in numbers and most likely do not represent the actual bills, so we can leave them out of inference analysis since these values are much less accurate than the rest of the data and might skew the model. We'd want to have those outliers back when training the prediction model though.   
6. The missing values are only for two columns: "Property value" (roughly 30% of the data) and "Lot size" (about 17% of the rows). We'll have to deal with those later when working on predictive modelling.   
7.The boxplots do appear to hint on different electric bills depending on the building type but lets let's group those units on whether they could be considered a house or an apartment. We'll exclude mobile homes, trailers, boats, RVs and vans from the analysis.   

     

```{r, , fig.width=6, fig.height=3, echo=F, include = T}
PUMS_2015$YBL<-revalue(PUMS_2015$YBL, c("1939 or earlier"=1939,"1950 to 1959"=1959, "1990 to 1999"=1999,"1960 to 1969"=1969, "2000 to 2004"=2004, "1970 to 1979"=1979,"1980 to 1989"=1989, "1940 to 1949"=1949))
PUMS_2015$YBL<-ifelse(PUMS_2015$YBL>2000, 1, 0)
PUMS_2015$HEATING<-ifelse(PUMS_2015$HFL=='Electricity', 1 , 0)
PUMS_2015$HOUSE<-ifelse(PUMS_2015$BLD=="One-family house detached" | PUMS_2015$BLD=="One-family house attached", 1 , 0)

PUMS_2015_inferential<-PUMS_2015[PUMS_2015$BLD!= 'Mobile home or trailer' & PUMS_2015$BLD!='Boat, RV, van, etc.', ]


data.elec<-subset(PUMS_2015_inferential, ELEP<500)

data.elec$HOUSING<-ifelse(data.elec$HOUSE==1,"Yes","No")

ggplot(data.elec, aes(factor(HOUSING), ELEP, fill = factor(HOUSING))) + 
  geom_boxplot()+theme_tufte()+theme(text = element_text(size=rel(3)),legend.position = "none")+xlab("Living in a house?")+ylab(colnames(PUMS_2015.nice[7]))
```

There appear to be some difference in electricity bills depending on whether the people live in a house. Now we'll build inferential model to test that hypothesis and estimate the differences. We'll adjust for the following variables:
1. Number of bedrooms (numerical).   
2. Number of occupants (numerical).   
3. Year the structure was first built (converted to categorical variable as before or after 2000 to account for the modern energy-efficient home designs).  
4. House heating fuel as electricity vs. other (categorical).    
5. Presence	of persons under 18 years in household (categorical).     
6. Interaction term for the number of bedrooms when electricity is used as a heating fuel.  

```{r, echo=F, include = F}
lm.inference.int<-lm(ELEP~NP+BDSP+HEATING+YBL+HOUSE+R18+HEATING:BDSP, data<-data.elec)
model<-lm.inference.int%>%tidy()
#kable(model)%>%
 # kable_styling(latex_options = c("striped","scale_down"), stripe_color = "gray!15", font_size = 7, full_width = TRUE)
summary(lm.inference.int)
```
First, let's check if our interaction term is justified. We can compare two models (with vs. without interaction) using extra SS F-test:

```{r, echo=FALSE, include=TRUE}
lm.inference.noint<-lm(ELEP~NP+BDSP+HEATING+YBL+HOUSE+R18, data<-data.elec)
anova(lm.inference.noint, lm.inference.int)
```
There is convincing evidence that our response variable is associated with at least one of interaction terms, (p = 1.51e-13, extra sum of squares F-test on 6 and 13675 degrees of freedom). We'll keep the interaction term in the model. 

Next, let's perform collinearity test between numeric columns and residual diagnostics to check the model satisfies regression assumptions:


```{r, include=TRUE, echo=FALSE}
ggplot(data.elec, aes(NP, BDSP))+geom_point()+theme_tufte()+geom_jitter()+labs(x="Number of occupants",y="Number of bedrooms")
```



```{r, fig.width=6, fig.height=3, include = T, echo=FALSE,  message=FALSE, warning=FALSE}
data.additional<- fortify(lm.inference.int, data = data.elec)
p1<-ggplot(data.additional, aes(x=.resid)) + geom_histogram()+xlab("Residuals")+ylab("Count")+theme_tufte()+theme(text = element_text(size=rel(3)))
p2<-ggplot(data.additional, aes(y=.resid, x=.fitted))+ geom_point()+xlab("Fitted Values")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
p3<-ggplot(data.additional, aes(y=.fitted, x=ELEP))+ geom_point()+xlab("Observed values")+ylab("Fitted Values")+theme_tufte()+theme(text = element_text(size=rel(3)))
p4<-ggplot(data.additional, aes(y=.resid, x=ELEP))+ geom_point()+xlab("Observed values")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
p5<-ggplot(data.additional, aes(y=.resid, x=NP))+ geom_point()+xlab("Number of Person Records per Household")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
p6<-ggplot(data.additional, aes(y=.resid, x=BDSP))+ geom_point()+xlab("Number of Bedrooms")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
grid.arrange(p1,p2,p3,p4,p5,p6, nrow=3, top = 'Checking Linear Regression Assumptions for the Original Function:')
```

There's not much collinearity between the numerical columns. As for the regression assumption, the histogram of the residuals is skewed to the right which violates the normality assumption, although linear regression models are typically robust to this violation, especially for the large sample sizes that we have here. Also, observed vs. fitted values doesn't show linear relationship between two variables. Let's transform the response with log operator and see if it helps with the residuals distribution:

```{r, fig.width=6, fig.height=3, include = T, echo=FALSE,  message=FALSE, warning=FALSE}
lm.inference.int<-lm(log(ELEP)~NP+BDSP+HEATING+YBL+HOUSE+R18, data<-data.elec)
data.additional<- fortify(lm.inference.int, data = data.elec)
p1<-ggplot(data.additional, aes(x=.resid)) + geom_histogram()+theme_tufte()+theme(text = element_text(size=rel(3)))+labs(x="Residuals", y="Count")
p2<-ggplot(data.additional, aes(y=.resid, x=.fitted))+ geom_point()+theme_tufte()+theme(text = element_text(size=rel(3)))+labs(x="Fitted Values", y="Residuals")
p3<-ggplot(data.additional, aes(y=.fitted, x=log(ELEP)))+ geom_point()+xlab("Observed values")+ylab("Fitted Values")+theme_tufte()+theme(text = element_text(size=rel(3)))
p4<-ggplot(data.additional, aes(y=.resid, x=log(ELEP)))+ geom_point()+xlab("Observed values")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
p5<-ggplot(data.additional, aes(y=.resid, x=NP))+ geom_point()+xlab("Number of Person Records per Household")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
p6<-ggplot(data.additional, aes(y=.resid, x=BDSP))+ geom_point()+xlab("Number of Bedrooms")+ylab("Residuals")+theme_tufte()+theme(text = element_text(size=rel(3)))
grid.arrange(p1,p2,p3,p4,p5,p6, nrow=3, top = 'Checking Linear Regression Assumptions for Log-transformed Function:')
```

There's still some skewedness in residuals distribution for very high and very low electricity bills but it's much better with log transform and should be good enough for the inference task. Also, the fitted vs. observed value correlation does appear to be linear.


```{r, echo=FALSE, include=TRUE}
lm.inference.int<-lm(log(ELEP)~NP+BDSP+HEATING+YBL+HOUSE+R18+HEATING:BDSP, data<-data.elec)
summary(lm.inference.int)
ci<-confint(lm.inference.int)
(ci)
```


```{r}
log.ci.convert<-function(n){(exp(n)-1) * 100}
(ci.low<-round(log.ci.convert(ci[6,1]),2))
(ci.high<-round(log.ci.convert(ci[6,2]),2))
(point.est<-round(log.ci.convert(lm.inference.int$coefficients[6]),2))
```

Since we are using the log-transformation, the coefficient only represented change of the ratio of electricity cost between house and apartment. Thus, we'd need to do some transformation of the response to get the confidence interval.

#### Results
There is convincing evidence that people living in houses in Oregon state in 2015 were on average paying more for electricity than those living in apartments (p-value < 2.2e-16), afteraccounting for the number of bedrooms, number of occupants in the household, whether the dwelling was built after 2000, whether it is using electricity for heating, for the presence	of persons under 18 years in household, and the number of bedrooms when electricity is used as a heating fuel. It is estimated that living in a house increased the mean electricity monthly cost by `r point.est`%. With 95% confidence, the mean increase in electricity monthly cost was between `r ci.low`% and `r ci.high`%. The model was based on Public Use Microdata Sample (PUMS) from the 2015 1-year survey in Oregon and only included households that have at least one person, pay for their electricity, and are not group accommodations. The inference shouldn't be extrapolated to the dwellings with very high electricity bills ($540 and above) and institutional/non-institutional group	quarters as those were excluded from the analysis.    

2. Prediction Problem: Create a model that could be used to predict electricity costs for a household in Oregon.

```{r}
data.na.VALP<- subset(PUMS_2015, is.na(VALP))
head(data.na.VALP)
summary(data.na.VALP)
table(data.na.VALP$TEN)

data.na.ACR<- subset(PUMS_2015, is.na(ACR))
head(data.na.ACR)
summary(data.na.ACR)
```


```{r}
kable(table(data.na.ACR$BLD))%>%
  kable_styling(latex_options = c("striped","scale_down"), stripe_color = "gray!15", font_size = 10, full_width = TRUE)
```

First, we'll clean up the data to get it ready for the modeling. Let's explore that VALP and ACR columns which had missing values. As expected, missing values in "Lot size" column are associated with people who don't live in houses. So we add those NA values as an extra factor to the column. NAs in "Property value" column might correspond to the properties which are either rented or occupied without payment or rent(30%). We can't find a suitable substitute for NAs in this case so we'll omit the whole column.
We'll also remove SERIALNO and TYPE columns as irrelevant ones and "Units in structure" and "House heating fuel" columns to their derivatives ("Does person live in a house?" and "Is electricity used as house heating fuel?"). We'll also convert R18 and R60 columns to numeric values.


```{r}
data.elec.predict<-PUMS_2015
data.elec.predict$R18<-ifelse(data.elec.predict$R18=='1 or more', 1, 0)
data.elec.predict$R60<-ifelse(data.elec.predict$R60=='1 or more', 1, 0)
data.elec.predict<-subset(data.elec.predict, select=-c(SERIALNO,TYPE,BLD,HFL,VALP))
data.elec.predict$ACR <- addNA(factor(data.elec.predict$ACR))
#pairs(select_if(data.elec.predict, is.numeric))
```

Let's split the data into train/validate/test (80%/10%/10%) and proceed with the modelling.

```{r}
set.seed(123) 
split = sample.split(data.elec.predict, SplitRatio = 0.8)
training_set = subset(data.elec.predict, split == TRUE)

training_set.named<-training_set
colnames(training_set.named)<-c("Records per housing record","Lot size","Number of bedrooms", "ELEP", "Yearly fuel cost (excl. gas and electricity)","Gas (monthly cost)", "Number of Rooms", "Tenure",  "Was structure built after 2000", "Under 18 years in household", "60 years and over in household", "Does person live in a house?","Is electricity used as house heating fuel?")

testing_set = subset(data.elec.predict, split == FALSE)

set.seed(321) 
split = sample.split(testing_set, SplitRatio = 0.5)
validation_set = subset(testing_set, split == TRUE)
test_set = subset(testing_set, split == FALSE)
```

```{r}
str(training_set.named)
```

We'll use log transformation as it showed to have a better alignment with the regression model assumptions. Let's perform a feature selection analysis on the training set using best subset regression techniques with BIC, CP and Adjusted R2 as the metrics of choice.

```{r, echo=FALSE, include=TRUE}
regfit.full<-regsubsets(log(ELEP)~.,training_set.named, nvmax = 20)
regfit.summary<-summary(regfit.full)
#regfit.summary

par(mfrow=c(2,3), mar=c(1,1,1,1),cex.axis = 0.5, cex.lab = 0.2, cex.main=0.5)
#plot(regfit.summary$rss, xlab="Number of Variables", ylab="RSS", type="l")
plot(regfit.summary$adjr2, xlab="Number of Variables", main="Adjusted RSq", type="l")
max.adjr2<-which.max(regfit.summary$adjr2)
points(max.adjr2, regfit.summary$adjr2[max.adjr2], col="red", cex=2, pch=20)

plot(regfit.summary$cp, xlab="Number of Variables", main ="Cp", type="l")
min.cp<-which.min(regfit.summary$cp) 
points(min.cp, regfit.summary$cp[min.cp], col="red", cex=2, pch=20)

min.bic<-which.min(regfit.summary$bic)
plot(regfit.summary$bic, xlab="Number of Variables", main="BIC", type="l")
points(min.bic, regfit.summary$bic[min.bic], col="red", cex=2, pch=20)
plot(regfit.full, scale="adjr2")
plot(regfit.full, scale="Cp")
plot(regfit.full, scale="bic")
```
Out of 16 variables and intercept, Adjusted Rsq chose 14 variables (excluding "Tenure rented" and "Tenure owned with mortgage and loan"). Cp would exclude "Under 18 years in household" along with two factors of TEN column. BIC model would also exclude "60 years and over in household" (12 variable model).


Now let's do a 10-fold cross-validation to check which one of the models produces the least error (MSE) on the validation set and hence is the best model.

```{r, warning=FALSE}
#We'll need to split TENR to its factor columns
training_set$TEN_O<-if_else(training_set$TEN=="Owned free and clear", 1, 0)
training_set$TEN_L<-if_else(training_set$TEN=="Owned with mortgage or loan", 1, 0)
training_set$TEN_R<-if_else(training_set$TEN=="Rented", 1, 0)
validation_set$TEN_O<-if_else(validation_set$TEN=="Owned free and clear", 1, 0)
validation_set$TEN_L<-if_else(validation_set$TEN=="Owned with mortgage or loan", 1, 0)
validation_set$TEN_R<-if_else(validation_set$TEN=="Rented", 1, 0)
test_set$TEN_O<-if_else(test_set$TEN=="Owned free and clear", 1, 0)
test_set$TEN_L<-if_else(test_set$TEN=="Owned with mortgage or loan", 1, 0)
test_set$TEN_R<-if_else(test_set$TEN=="Rented", 1, 0)
data.elec.predict$TEN_O<-if_else(data.elec.predict$TEN=="Owned free and clear", 1, 0)
data.elec.predict$TEN_L<-if_else(data.elec.predict$TEN=="Owned with mortgage or loan", 1, 0)
data.elec.predict$TEN_R<-if_else(data.elec.predict$TEN=="Rented", 1, 0)

cv.new.error<-rep(0,3)

glm.fit.AdjR2 = glm(formula = log(ELEP) ~ .-TEN_L-TEN_R-TEN, data = training_set)
cv.new.error[1]<-cv.glm(validation_set, glm.fit.AdjR2, K=10)$delta[1]
glm.fit.Cp = glm(formula = log(ELEP) ~ .-R18-TEN_L-TEN_R-TEN, data = training_set)
cv.new.error[2]<-cv.glm(validation_set, glm.fit.Cp, K=10)$delta[1]
glm.fit.BIC = glm(formula = log(ELEP) ~ .-R18-R60-TEN_L-TEN_R-TEN, data = training_set)
cv.new.error[3]<-cv.glm(validation_set, glm.fit.BIC, K=10)$delta[1]
```


```{r, echo = FALSE, include=TRUE}
best.models <- data.frame(c('Adjusted R-squared', 'Cp','BIC'),c('14 variables', '13 variables','12 variables'),cv.new.error)
colnames(best.models)<- c('Metrics used for variable selection', 'Number of variables in the model', 'CV Error on testing data')
kable(best.models, format="latex", booktabs=TRUE) %>%
  kable_styling(latex_options = c("striped","HOLD_position"), stripe_color = "gray!15")
```

The model devised with CP metrics shows the smallest CV error.Let's use the full dataset to train the final model and predict electricity costs for a household in Oregon.


```{r}
data.elec.predict<-rbind(training_set, validation_set)
data.elec.predict.copy<-data.elec.predict
colnames(data.elec.predict)<-c("Records per housing record","Lot size","Number of bedrooms", "ELEP", "Yearly fuel cost (excl. gas and electricity)","Gas (monthly cost)", "Number of Rooms", "TEN",  "Was structure built after 2000", "R18", "R60","Is electricity used as house heating fuel?", "Does person live in a house?", "Tenure owned and free and clear","TEN_L", "TEN_R")
best.model.prediction<-lm(formula = log(ELEP) ~ . - R18 - TEN_L- TEN_R - TEN, 
    data = data.elec.predict)
summary(best.model.prediction)
coefficients(best.model.prediction)
prediction.formula<-paste0("log(ELEP)~ ", round(as.vector(coefficients(best.model.prediction)[1]), 5))
for (i in 2:length(coefficients(best.model.prediction))){
      prediction.formula<-paste0(prediction.formula,"+",round(as.vector(coefficients(best.model.prediction)[i]), 5),"*", names(coefficients(best.model.prediction)[i]))
}
j<-0
```


```{r, echo=FALSE, include=TRUE}
while (j<nchar(prediction.formula)){
  print(substr(prediction.formula, j+1, j+90))
  j<-j+90}
```

Finally, we'll run a prediction on the test dataset and calculate its RMSE.

```{r}
colnames(test_set)<-colnames(data.elec.predict)
predictions <- best.model.prediction %>% predict(test_set, interval="prediction", level=0.90)
predictions.ci <- best.model.prediction %>% predict(test_set, interval="confidence", level=0.90)
colnames(predictions.ci)<-c("fit", "lower.ci","upper.ci")
# Model performance
# (a) Prediction error, RMSE
(pred.error.test<-rmse(predictions, log(test_set$ELEP)))
```



```{r, include=TRUE, echo=FALSE}
pred.interval<-cbind(test_set, predictions, predictions.ci[,2],predictions.ci[,3])
ggplot(pred.interval, aes(x=exp(fit)))+geom_point(aes(y=ELEP), size=1)+geom_point(aes(y=exp(fit)), size=1)+geom_point(aes(y=exp(upr)), col="red")+geom_point(aes(y=exp(lwr)), col="red")+geom_point(aes(y=exp(predictions.ci[,2])), col="blue")+geom_point(aes(y=exp(predictions.ci[,3])), col="blue")+theme_tufte()+labs(x="Predicted Electricity Bill", y="Actual Electricity Bill", title="90% Confidence vs. Prediction Intervals")
```



#### Results

Here we have a model predicting monthly electricity cost for a household in Oregon. The cross-validation estimate of prediction error is `r round(cv.new.error[2],4)` and RMSE for test set prediction is `r round(pred.error.test,4)` on a log scale. The model is based on Public Use Microdata Sample (PUMS) from the 2015 1-year survey in Oregon and only includes households that have at least one person, pay for their electricity, and are not group accommodation. The property value column was excluded from the analysis due to missing 30% of the cases which might have affected its accuracy. Adjusted R-squared value for the model is only 0.274 and RMSE is quite large which might indicate that a linear regression model is not the best choice for the data at hand.    




## Conclusion
Below we tabulated the two approaches employed for the analysis above: drawing inferences vs. making predicitons.


```{r, echo = FALSE, include=TRUE}
discussion<-as.data.frame(matrix(nrow=9,ncol=2))
colnames(discussion)<-c('Explanatory Model', 'Predictive Model')
discussion[1,1]<-c('Primary goal is answering the question on difference in electricity bill for houses vs. apartments.')
discussion[1,2]<-c('Primary goal is having an accurate estimate of monthly electricity costs as well an estimate for the accuracy of the model.')

discussion[2,1]<-c('The strategy is to pick only those variables and interactions which are meaningful/needed for model interpretation (variables of direct concern). Limit	the	number	of	models	you	compare	and	choose	the	model	to	answer	the	scientific question	you	care	about. 	The	more	models	you	compare	the	more	likely	it	is	that	you	will	find	a	possibly	spurious	relationship')
discussion[2,2]<-c('The strategy is to choose the set of variables that will lead to a better prediction performance. If	your	goal	is	to	make	predictions	of	the	response	(or	mean response),	it	doesn’t	hurt	to	fit	and	compare	a	large	number of	models,	but	you	should	use	some	objective	way	of selecting	the	final	model (e.g. smallest AIC, smallest BIC, lowest prediction error on independent set of data)')


discussion[3,1]<-c('No concern for missing values and problematic columns as long as they are not in the selected variables.')
discussion[3,2]<-c('Need to clean up the data (missing values etc.) prior to selection.')

discussion[4,1]<-c('Need to avoid complicating the model interpretation with too much column transformations.')
discussion[4,2]<-c('All column transformations and feature engineering are allowed as long as they improve the accuracy of prediction.')

discussion[5,1]<-c('Cannot trust inference after variable selection (using the same data twice). Otherwise, we will be getting small p-values from the sample becuase we selected for those small p-values. That is, the process of variable selection invalidates the properties of the p-values and confidence intervals.')
discussion[5,2]<-c('Variable selection is the major step in model optimization.')

discussion[6,1]<-c('Typically selecting between only a couple of models (e.g. one with and without interactions) to avoid model selection bias.')
discussion[6,2]<-c('Comparing lots of models with best subset selection techniques and crossvalidation on independent dataset.')

discussion[7,1]<-c('Have to check for the regression assumptions by residual diagnostics and multicollinearity, Using log transformtion if needed.')
discussion[7,2]<-c('For prediction models adjusting for multicollinearity is not important unless it decreases the prediction error. Same is true for checking regression asumptions and transformations of variables.')

discussion[8,1]<-c('The explanatory variables for the model need to have significant slope coefficients.')
discussion[8,2]<-c('Did not care much if the coefficients for my model were statistically significant.')

discussion[9,1]<-c('Interested in p-values for the model and confidence interval for the variable of interest.')
discussion[9,2]<-c('Interested in prediction formula with coefficient and the prediction error.')

discussion[10,1]<-c('Violation of regression assumptions and collinearity as the main challenges.')
discussion[10,2]<-c('Low adjusted R-squared, poor model fit or overfitting as the main challenges.')

```

\renewcommand{\arraystretch}{4}
```{r, echo = FALSE, include=TRUE}
kable(discussion, format="latex", booktabs=TRUE) %>%
  kable_styling(latex_options = c("striped","scale_down"), stripe_color = "gray!15", font_size = 7, full_width = TRUE)
```


# R code appendix  

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=TRUE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=I(60)), tidy=TRUE)
```
